---
title: "Ordination"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Setup

### libraries

```{r results = "hold"}
tellme <- function(name){print(paste0("Package ", name, " version: ", packageVersion(name)))}

library(tidyr); tellme("tidyr")
suppressPackageStartupMessages(library(dplyr)); tellme("dplyr")
library(ggplot2); tellme("ggplot2")
library(ggrepel); tellme("ggrepel")
```
### Colors

A named vector of color vectors.  Most of these match the previous paper. 
```{r, fig.height=4, fig.width=2, echo=FALSE, include=FALSE}
themeFile = "../../input/themes.R"
if (file.exists(themeFile)){
    source(themeFile)
    myColorPalette = getColorPalette()
    displayColorPalette(myColorPalette)
}else{
    myColorPalette = c(
        # timepoint
        nonED = "#1B9E77",
        HC = "#1B9E77",
        T1 = "#D95F02",
        AN.AD = "#D95F02",
        T2 = "#7570B3",
        AN.DS = "#7570B3",
        # site
        Denver = "#1F78B4",
        ACUTE = "#1F78B4",
        UNC = "#F0027F",
        CEED = "#F0027F",
        FARGO = "coral",
        # subgroups indicating site AND group/timepoint
        UNC_HC = "#B2DF8A",
        Denver_HC = "#33A02C",
        UNC_T1 = "#FDBF6F",
        UNC_T2 = "#FF7F00",
        Denver_T1 = "#CAB2D6",
        Denver_T2 = "#6A3D9A"
    )
    # plot.new(); legend(legend=names(myColorPalette), 
    #                    col=myColorPalette, bty="n", x="center", pch=15, cex=2)
    
    colorKey = data.frame(color=myColorPalette, key=names(myColorPalette))
    # ggplot(data=colorKey, aes(x=1, y = 1:nrow(colorKey), fill=key, label=key)) +
    #     geom_tile() +
    #     scale_fill_manual(values = myColorPalette) +
    #     theme_void()+
    #     theme(legend.position="none") + 
    #     geom_text()
}

# Use the theme_classic() ggplot theme unless otherwise indicated.
theme_set(theme_classic())
```

Pick a taxonomic level. Take an argument if one is given.
```{r}
args = commandArgs(trailingOnly=TRUE)
if (length(args) > 0){
    taxaLevel = args[1]
}else{
    taxaLevel = "family"
}
```


Direct output
```{r}
outDir = file.path("..", "output", taxaLevel)
suppressWarnings( dir.create(outDir, recursive = T) )
```

Output will be saved to: ``r outDir``

# Main

## Read counts data

```{r}
input=c(
    domain = "../../input/counts-tables/filtered-counts_level-1_domain.csv",
    phylum = "../../input/counts-tables/filtered-counts_level-2_phylum.csv",
    class = "../../input/counts-tables/filtered-counts_level-3_class.csv",
    order = "../../input/counts-tables/filtered-counts_level-4_order.csv",
    family = "../../input/counts-tables/filtered-counts_level-5_family.csv",
    genus = "../../input/counts-tables/filtered-counts_level-6_genus.csv",
    species = "../../input/counts-tables/filtered-counts_level-7_species.csv")

# taxaLevel = "class"
countsFile = input[taxaLevel]
```

We are looking at the ``r taxaLevel`` level, so we will read file ``r countsFile``.

```{r}
countsAndMeta = read.csv(countsFile)
dim(countsAndMeta)
```

Filter out unrelated samples
```{r}
# already done, but double check
if (any(countsAndMeta$Person != "Kylie")){
    message("Filtering out unrelated samples.")
    countsAndMeta = countsAndMeta %>% filter(Person == "Kylie")
}else{
    message("No unrelated samples.")
}
```

```{r echo=FALSE, include=FALSE}
# Make a table with all the metadata that came with the artifact.
# No real need to save this.

# qzaMeta = countsAndMetaALL %>% select(id.orig, StudyID, Timepoint, Person)
# write.table(qzaMeta, file=file.path(outDir, paste0(taxaLevel, "_artifact_metatdata.txt")), 
#             quote=F, sep="\t", row.names = F)
```

This data includes some metadata.
```{r}
metaCols = c("id.orig", "BarcodeNumber", "BarcodeSequenceFull", "BarcodeSequence", "LinkerPrimerSequence", "StudyID", "Timepoint", "Person", "index")
```

All the other columns should look like taxa names. Look at the first few characters. How often does that prefix appear?
```{r}
dataCols = setdiff(colnames(countsAndMeta), metaCols)
table(substr(dataCols, 0, 11))
```


Split the meta data and counts
```{r}
metaCols = c("id.orig", "BarcodeNumber", "BarcodeSequenceFull", "BarcodeSequence", "LinkerPrimerSequence", "StudyID", "Timepoint", "Person", "index")
counts = countsAndMeta %>% select(-all_of(metaCols))
row.names(counts) = countsAndMeta$id.orig
key = countsAndMeta %>% select(id.orig, StudyID, Timepoint)
```

### normalize counts

Normalize counts.
```{r}
lognorm <- function(table, log10.do = TRUE){
  # table - a table with rows for samples and columns for features
  #         samples names are row names.
  #         feature names are column names.
  #         the table values are all numeric
  sampleSums = rowSums(table)
  meanSampleDepth = mean(sampleSums)
  sampleProportion = t( apply(table, 1, function(row) row / sum(row)) )
  t2 = sampleProportion * meanSampleDepth
  if (log10.do) t3 = log10( t2 + 1 )
  else t3 = t2 + 1
  return( t3 )
}

# counts.norm = lognorm(counts)
```

### Shorten names

Rename taxa in counts file.  These names are very long.  Give them shorter names that will be easy to read in tables and look nice in figures.

Function to display taxa names.
```{r}
shortenTaxonName = function(name, level=NULL){
    name2 = sub(".d__", "<<splitHere>>d__", name)
    name2 = sub(".p__", "<<splitHere>>p__", name2)
    name2 = sub(".c__", "<<splitHere>>c__", name2)
    name2 = sub(".o__", "<<splitHere>>o__", name2)
    name2 = sub(".f__", "<<splitHere>>f__", name2)
    name2 = sub(".g__", "<<splitHere>>g__", name2)
    name2 = sub(".s__", "<<splitHere>>s__", name2)
    parts = unlist(strsplit(name2, split="<<splitHere>>", fixed=T))
    bits = strsplit(parts, split="__")
    levelName = sapply(bits, "[", 1)
    taxaName = sapply(bits, "[", 2)
    names(taxaName) = levelName
    
    lastBit = taxaName[length(taxaName)]
    nextLastPart = parts[length(parts) - 1]
    
    if (is.na(lastBit) | lastBit == "." | lastBit == ""){
        result = nextLastPart
    }else if (lastBit == "uncultured_bacterium"){
        result = paste0(nextLastPart, " (UB*)")
    }else{
        result = lastBit
    }
    
    if (!is.null(level)){
        result = paste(level, result, sep=": ")
    }
    return(result)
}

displayTaxonName <- function(name, level=NULL, maxchar=50){
    if (!is.na(maxchar) & nchar(name) > maxchar){
        ending = substr(name, start=nchar(name)-maxchar-2, stop=nchar(name))
        name = paste0("...", ending)
    }
    if (!is.null(level)){
        name = paste(level, name, sep=": ")
    }
    return(name)
}
```

If the new names are not all unique, then append an arbitrary number to them.  Either way, save a key of original taxa names and shortened taxa names.
```{r}
originalName = colnames(counts)

shortName = sapply(originalName, shortenTaxonName)

if (length(unique(shortName)) < ncol(counts)){
    shortName = paste0(shortName, "[", 1:ncol(counts), "]")
}
colnames(counts) = shortName

nameKey = data.frame(shortName=shortName, 
                       originalName=originalName)

write.table(nameKey, 
            file = file.path(outDir, "TaxonNameKey.txt"), 
            quote=F, row.names=F, sep="\t")
```


## Read meta data.

Read the original metadata file.
```{r}
meta = read.delim("../../input/meta/ANIGMA-metadata.txt") %>%
  select(PARTICIPANT.ID, LOCATION, TIMEPOINT, AGE, SUBTYPE, BMI, 
         STAI_Y1, STAI_Y2, STAI_TOTAL, PSS, DAYS_TREAT, Weight_kg, DNA.ID, DUR_ILLNESS_YRS) %>%
  filter(DNA.ID %in% countsAndMeta$id.orig)
row.names(meta) = meta$DNA.ID
dim(meta)
```

Check metadata.
```{r}
if (any(!is.na(meta$DAYS_TREAT[meta$TIMEPOINT=="HC"]))){
    message("For any Healthy Control participant, the DAYS_TREAT should be NA (not 0).")
    meta$DAYS_TREAT[meta$TIMEPOINT=="HC"] = NA
}else{
    message("Ah, good, for Healthy Control participants, the DAYS_TREAT is NA (not 0).")
}
```

Check that the meta data that came with the qiime artifact matches up to the main project meta data.
```{r}
m = merge(meta, key, by.x="DNA.ID", by.y="id.orig")
table(m$PARTICIPANT.ID == m$StudyID)
table(m$TIMEPOINT == m$Timepoint)
```

The metadata from the qiime artifact and the metadata for the project match up. There is one that has an error in the qiime artifact metadata. That's fine. We'll use the project metadata moving forward.
```{r}
m[which(m$TIMEPOINT != m$Timepoint), c("PARTICIPANT.ID", "TIMEPOINT", "BMI", "Timepoint")]
```

### Feature engineering

For subtype, we are only interested in the AN cases. HC is not really an AN subtype. And we need to split samples between T1 and T2 so we are not double-dipping with each paired sample.
```{r}
meta$SUBTYPE.AN = meta$SUBTYPE
meta$SUBTYPE.AN[meta$SUBTYPE.AN=="HC"] = NA
```

The anorexia signal is what we would see by comparing T1 to HC or T2 to HC, which is the "TIMEPOINT" variable. Lets call it the ANOREXIA variable so that its clearer in the summaries.
```{r}
meta$ANOREXIA = meta$TIMEPOINT
```


### Previous feature engineering

Read the diffs metadata - data where diffs have been calculated.
```{r}
prevModule = dir(path="../..", pattern="Participant_Metadata", full.names = T)
diffsFile = file.path(prevModule, "output", "ANIGMA-metadata_by_AN_participant.txt")
diffs = read.delim(diffsFile)
row.names(diffs) = diffs$PARTICIPANT.ID
dim(diffs)
```

Read in modified metadata from: ``r diffsFile``

This table has ``r nrow(diffs)`` rows and ``r ncol(diffs)`` columns.

Match this data withe corresponding DNA ID for T1 and T2.
```{r}
diffKey = meta %>% 
    select(PARTICIPANT.ID, TIMEPOINT, DNA.ID) %>%
    pivot_wider(id_cols = PARTICIPANT.ID, names_from = TIMEPOINT, values_from = DNA.ID, names_prefix = "DNA.ID.") %>%
    select(-DNA.ID.HC)
diffs = merge(diffs, diffKey, all.x=T, by="PARTICIPANT.ID")
```


## merge

<!-- Merge the metadata and the normalized counts data. -->
<!-- ```{r} -->
<!-- data = merge(counts.norm, meta, by.x=0, by.y="DNA.ID") -->
<!-- dim(data) -->
<!-- ``` -->

<!-- Make a scrambled version. We can use this to verify that we see null results, and our tests are not inherently flawed. -->
<!-- ```{r} -->
<!-- scram.counts = counts.norm -->
<!-- row.names(scram.counts) = sample(row.names(scram.counts), size=nrow(scram.counts), replace = FALSE) -->
<!-- scrambled.data = merge(scram.counts, meta, by.x=0, by.y="DNA.ID") -->
<!-- rm(scram.counts) -->
<!-- ``` -->

<!-- Merge the T1 counts data to the per-participant differences data. -->
<!-- ```{r} -->
<!-- diffData1 = merge(counts.norm, diffs, by.x=0, by.y="DNA.ID.T1") -->
<!-- diffData2 = merge(counts.norm, diffs, by.x=0, by.y="DNA.ID.T2") -->
<!-- ``` -->


<!-- Make sure Patient ID is not treated as a numerical value. -->
<!-- ```{r} -->
<!-- data$PARTICIPANT.ID = as.character(data$PARTICIPANT.ID) -->
<!-- scrambled.data$PARTICIPANT.ID = as.character(scrambled.data$PARTICIPANT.ID) -->
<!-- ``` -->

# Test

## Phyloseq

```{r}
library(phyloseq)

ps <- phyloseq(otu_table(counts, taxa_are_rows=FALSE), 
                 sample_data(meta))

ps.prop <- transform_sample_counts(ps, function(otu) otu/sum(otu))

ord <- ordinate(ps.prop, method="PCoA", distance="bray") 
```

```{r}
po12 = plot_ordination(ps.prop, ord, color="TIMEPOINT") + 
    scale_colour_manual(values = myColorPalette) +
    theme_minimal()
show(po12)

po34 = plot_ordination(ps.prop, ord, color="TIMEPOINT", axes = c(3,4)) + 
    scale_colour_manual(values = myColorPalette) +
    theme_minimal()
show(po34)
```

