---
title: "Reshape Metadata"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Settup

### Libraries

```{r results = "hold"}
tellme <- function(name){print(paste0("Package ", name, " version: ", packageVersion(name)))}

library(tidyr); tellme("tidyr")
suppressPackageStartupMessages(library(dplyr)); tellme("dplyr")
library(ggplot2); tellme("ggplot2")
library(ggrepel); tellme("ggrepel")
```

Direct output.
```{r echo=FALSE}
outdir = "../output"
suppressWarnings( dir.create(outdir) )
```

Output will be saved to: ``r outdir``

### Read data

Read data.
```{r}
meta = read.delim("../../input/meta/ANIGMA-metadata.txt") %>%
  select(PARTICIPANT.ID, LOCATION, TIMEPOINT, AGE, SUBTYPE, BMI, 
         STAI_Y1, STAI_Y2, STAI_TOTAL, PSS, DAYS_TREAT, Weight_kg)
dim(meta)
```

Get the per-particpant metadata.
```{r}
pipeline = "../../"
prevModule = dir(path=pipeline, pattern="Participant_Metadata", full.names = TRUE)
fileName = file.path(prevModule, "output", "ANIGMA-metadata_by_AN_participant.txt")
diffs = read.delim(fileName)
```

Read in data from file: ``r fileName``

This table has ``r nrow(diffs)`` rows and ``r ncol(diffs)`` columns.


### Colors

A named vector of color vectors.  Most of these match the previous paper. 
```{r, fig.height=4, fig.width=2, echo=FALSE, include=FALSE}
themeFile = "../../input/themes.R"
if (file.exists(themeFile)){
    source(themeFile)
    myColorPalette = getColorPalette()
    displayColorPalette(myColorPalette)
}else{
    myColorPalette = c(
        # timepoint
        nonED = "#1B9E77",
        HC = "#1B9E77",
        T1 = "#D95F02",
        AN.AD = "#D95F02",
        T2 = "#7570B3",
        AN.DS = "#7570B3",
        # site
        Denver = "#1F78B4",
        ACUTE = "#1F78B4",
        UNC = "#F0027F",
        CEED = "#F0027F",
        FARGO = "coral",
        # subgroups indicating site AND group/timepoint
        UNC_HC = "#B2DF8A",
        Denver_HC = "#33A02C",
        UNC_T1 = "#FDBF6F",
        UNC_T2 = "#FF7F00",
        Denver_T1 = "#CAB2D6",
        Denver_T2 = "#6A3D9A"
    )
    # plot.new(); legend(legend=names(myColorPalette), 
    #                    col=myColorPalette, bty="n", x="center", pch=15, cex=2)
    
    colorKey = data.frame(color=myColorPalette, key=names(myColorPalette))
    ggplot(data=colorKey, aes(x=1, y = 1:nrow(colorKey), fill=key, label=key)) +
        geom_tile() +
        scale_fill_manual(values = myColorPalette) +
        theme_void()+
        theme(legend.position="none") + 
        geom_text()
}
```

Use the theme_classic() ggplot theme unless otherwise indicated.
```{r}
theme_set(theme_classic())
```


### Remove/modify individual data.

Note: 469017 and 469019 are the outliers (same as before!) who lost weight during their stay, this was attributed to medication they had been taking ahead of their stay that lead to inflated weight values on arrival.  So the T1 weight (and by extension bmi) is not a valid measure here.

```{r}
meta[meta$PARTICIPANT.ID=="469017", "BMI"] = NA
meta[meta$PARTICIPANT.ID=="469019", "BMI"] = NA
```

Note: 469021 has a T2 PSS score of 0, which we believe is a data entry error.
Likewise for patient 469101, they have a pss score of 0 at T2.
```{r}
meta[meta$PARTICIPANT.ID=="469021" & meta$TIMEPOINT=="T2", "PSS"] = NA
meta[meta$PARTICIPANT.ID=="469101" & meta$TIMEPOINT=="T1", "PSS"] = NA
```

## MAIN

### BMI

Body mass index (BMI) is a central metic of outcome for these patients.  A normal range is 18.5 to 25.
```{r}
bmi.cat.breaks = c(18.5, 25)
assign.pss.category <- function(x){
  result = rep(NA, length(x))
  result[x < 18.5] = 1
  result[x >= 18.5 & x < 25] = 2
  result[x >= 25 ] = 3
  return(result)
}
```

#### BMI boxplot

```{r fig.height=3, fig.width=5}
figureLabels = c("Non-ED","AN-AD","AN-DS")

bmiP = ggplot(data=meta,
                  aes(x=TIMEPOINT, y=BMI, col=TIMEPOINT)) +
    scale_colour_manual(values = myColorPalette,
                        labels=figureLabels, name="Group") +
    geom_hline(yintercept=bmi.cat.breaks, color="gray80", size=1) +
    geom_boxplot(outlier.alpha = 0) +
    geom_jitter(width = .07, height = 0, size=.5) +
    xlab("") +
    annotate("text", y=mean(c(18.5, 25)), x=.5, label="Normal",
             color="gray70", angle=90, size=3) +
    scale_x_discrete(labels=figureLabels) +
    ggtitle("Body Mass Index") + 
    theme_classic() 
bmiP

bmiP1 = bmiP + theme(legend.position="none")
```

```{r}
bmi.plot.file = file.path(outdir, "bmi-boxplot.png")
ggsave(bmiP1, filename = bmi.plot.file, height = 2.5, width = 3)
```

Saved: ``r bmi.plot.file``

#### BMI by site

Separated by site.
```{r fig.height=3, fig.width=7}
bmiP2 = bmiP1 + facet_wrap("LOCATION")
bmiP2
```

```{r}
bmi.plot.file.site = file.path(outdir, "bmi-boxplot_by-site.png")
ggsave(bmiP2, filename = bmi.plot.file.site, height = 3, width = 7)
```

Saved: ``r bmi.plot.file.site``

#### BMI difference

```{r}
bmi.diff = ggplot(data=diffs, aes(x=BMI.diff)) +
    geom_vline(xintercept=0, color="gray5", size=1) + 
    geom_histogram(binwidth = .5, fill="coral", col="gray5") + 
    ggtitle("BMI difference") + 
    xlab("")
bmi.diff
```


```{r}
add_overhead_boxplot = function(gghistogram, xvals, yish){
    xvals = na.omit(xvals)
    gghistogram + 
        geom_rect(ymin=yish-1, ymax = Inf, xmin=-Inf, xmax=Inf, fill="white") +
        geom_segment(x=min(xvals), xend=max(xvals), y=yish, yend=yish) +
        geom_rect(xmin=quantile(xvals, .25), xmax=mean(xvals), 
                  ymin=yish-.5, ymax=yish+.5, fill="white", color="gray5") +
        geom_rect(xmin=mean(xvals), xmax=quantile(xvals, .75), 
                  ymin=yish-.5, ymax=yish+.5, fill="white", color="gray5") + 
        ylim(min=NA, max=yish+1)
}

add_overhead_boxplot(bmi.diff, xvals=diffs$BMI.diff, yish=18.5)
```


```{r}
bmi.diff.file = file.path(outdir, "bmi-difference.png")
ggsave(filename = bmi.diff.file, height = 3, width = 7)
```

Saved: ``r bmi.diff.file``



#### BMI difference by site
```{r}
ggplot(data=diffs, aes(x=BMI.diff, fill=LOCATION)) +
    scale_fill_manual(values = myColorPalette) +
    geom_vline(xintercept=0, color="gray5", size=1) + 
    geom_histogram(binwidth = .5) + facet_wrap("LOCATION")
```

```{r}
bmi.diff.file.site = file.path(outdir, "bmi-difference_by-site.png")
ggsave(filename = bmi.diff.file.site, height = 3, width = 7)
```

Saved: ``r bmi.diff.file.site``

#### BMI gain per day

```{r}
bmi.pday = ggplot(data=diffs, aes(x=BMI.gain.per.day)) +
    geom_vline(xintercept=0, color="gray5", size=1) + 
    geom_histogram(fill="purple", col="gray5") + 
    ggtitle("average daily BMI gain") + 
    xlab("")
bmi.pday
```

```{r}
bmi.pday.file = file.path(outdir, "bmi-mean-daily-gain.png")
ggsave(filename = bmi.pday.file, height = 3, width = 7)
```

Saved: ``r bmi.pday.file``

```{r}
bmi.pday + 
    scale_fill_manual(values = myColorPalette) +
    facet_wrap("LOCATION")
```



### PSS

Perceived Stress Scale (PSS) is a survey-based metric of patient stress.

Low is 0-13.
Moderate is 14-26.
High is 27-40.
```{r}
pss.cat.breaks = c(0, 13.5, 26.5, 40)
assign.pss.category <- function(x){
  result = rep(NA, length(x))
  result[x < 14] = 1
  result[x >= 14 & x < 27] = 2
  result[x >=27 ] = 3
  return(result)
}
```

#### PSS boxplot

Previously, Kylie showed this data with boxplots.
```{r fig.height=2.5, fig.width=4}
add_pss_labels = function(ggPlotObj){
    gg2 = ggPlotObj + 
    annotate("text", y=c(mean(pss.cat.breaks[c(1,2)]),
                         mean(pss.cat.breaks[c(2,3)]),
                         mean(pss.cat.breaks[c(3,4)])),
             x=rep(0.5,3),
             label=c("Low", "Moderate", "High"),
             color="gray70", angle=90, size=2.5)
    return(gg2)
}
p0 = ggplot(data=meta,
            aes(x=TIMEPOINT, y=PSS, col=TIMEPOINT)) +
    scale_colour_manual(values = myColorPalette,
                        labels=figureLabels, name="Group") +
    geom_hline(yintercept=pss.cat.breaks[c(2,3)], color="gray80", size=1.5) +
    geom_hline(yintercept=pss.cat.breaks[c(1,4)], color="gray70", size=1) +
    geom_boxplot(outlier.alpha = 0) +
    geom_jitter(width = .07, height = 0, size=.5) +
    scale_x_discrete(labels=figureLabels, name="") +
    ggtitle("Percieved Stress Scale")  +
    theme_classic() 
p1 = add_pss_labels(p0)
p1
```

```{r}
p2 = add_pss_labels(p0) + theme(legend.position="none")
pss.plot.file = file.path(outdir, "pss-boxplot.png")
ggsave(filename = pss.plot.file, height = 2.5, width = 3)
```

Saved: ``r pss.plot.file``


#### PSS by site

```{r}
p0
p0 + facet_wrap("LOCATION")
```

```{r}
pss.plot.site = file.path(outdir, "pss-boxplot_by-site.png")
ggsave(p0 + 
           facet_wrap("LOCATION") + 
           theme(legend.position="none"), 
       filename = pss.plot.site, height = 3, width = 7)
```

Saved: ``r pss.plot.site``

#### PSS difference

```{r}
pss.diff = ggplot(data=diffs, aes(x=PSS.diff)) +
    geom_vline(xintercept=0, color="gray5", size=1) + 
    geom_histogram(binwidth = 3, fill="dodgerblue", col="gray5") +
    ggtitle("PSS difference") +
    xlab("") +
    theme_classic()
pss.diff
```

```{r}
add_overhead_boxplot(pss.diff, xvals=diffs$PSS.diff, yish=17)
```

```{r}
pss.diff.file = file.path(outdir, "pss-difference.png")
ggsave(filename = pss.diff.file, height = 3, width = 7)
```

Saved: ``r pss.diff.file``

#### PSS difference by site

```{r}
ggplot(data=diffs, aes(x=PSS.diff, fill=LOCATION)) +
    geom_vline(xintercept=0, color="gray5", size=1) + 
    geom_histogram(binwidth = 3, col="gray5")  + 
    scale_fill_manual(values = myColorPalette) +
    theme(legend.position="none") +
    ggtitle("PSS difference") +
    xlab("") +
    facet_wrap("LOCATION")
```

```{r}
pss.diff.file.site = file.path(outdir, "pss-difference_by-site.png")
ggsave(filename = pss.diff.file.site, height = 3, width = 7)
```

Saved: ``r pss.diff.file.site``

### STAI

The State Trait Anxiety Index (STAI) is represented as Y1 and Y2, or as a total. The Y1 is State Anxiety component. The Y2 is the Trait anxiety component.

No or low anxiety: 40 - 74
Moderate anxiety: 75 - 88
High anxiety: 89 - 160

(Kylie may modify that scale later)

```{r}
stai.tot.breaks = c(74.5, 88.5)

assign.tot.Category = function(x){
  result = rep(NA, length(x))
  result[x < 75] = 1
  result[x >= 75 & x < 89] = 2
  result[x >=89 ] = 3
  return(result)
}
```

#### STAI boxplot

```{r}
staiDF = meta %>% 
    select(TIMEPOINT, STAI_Y1, STAI_Y2, STAI_TOTAL, LOCATION) %>%
    pivot_longer(cols=c(-TIMEPOINT, -LOCATION), names_to = "category", values_to = "stai")
```


```{r fig.height = 3, fig.width = 7}
ggplot(data=staiDF, aes(group_by=TIMEPOINT, y=stai, col=TIMEPOINT, x=category)) +
    scale_colour_manual(values = myColorPalette,
                        labels=figureLabels, name="Group") +
    # theme(legend.position="none") +
    # lines
    geom_hline(yintercept=seq(0,160,25), color="gray90", size=0.4) +
    # range labels
    geom_segment(x=-Inf, xend=1.5, y=74.5, yend=74.5, col="gray40") +
    geom_segment(x=-Inf, xend=1.5, y=88.5, yend=88.5, col="gray40") +
    geom_text(x=.42, y=mean(stai.tot.breaks), label="moderate", col="gray40", hjust="inward", size=2) +
    geom_text(x=.42, y=stai.tot.breaks[1], label="low", col="gray40", vjust = 1.2, hjust="inward") +
    geom_text(x=.42, y=stai.tot.breaks[2], label="high", col="gray40", vjust = -.2, hjust="inward") +
    # xlim(min=-.5, max=NA) +
    # data
    ggtitle("State Trait Anxiety Index") +
    geom_boxplot(outlier.alpha = 0) +
    geom_jitter(position=position_jitterdodge(jitter.width = .09, jitter.height = 0), size=.5) + 
    xlab("") + ylab("Anxiety Index") +
    theme_classic()

```


```{r}
stai.boxplot.file = file.path(outdir, "stai-boxplot.png")
ggsave(filename = stai.boxplot.file, height = 3, width = 7)
```

Saved: ``r stai.boxplot.file``

#### STAI differences

Histograms of differences in a grid to give the histogram of T2-T1 for each location and for each category (Y1, Y2, Total).

```{r}
staiDiff = diffs %>% 
    select(PARTICIPANT.ID, STAI_Y1.diff, STAI_Y2.diff, STAI_TOTAL.diff, LOCATION) %>%
    pivot_longer(cols=c(-PARTICIPANT.ID, -LOCATION), 
                 names_to = "category", values_to = "difference") %>%
    mutate(category = gsub(".diff", "", category))
```

```{r}
ggplot(data=staiDiff, aes(x=difference)) +
    geom_vline(xintercept=0, color="gray5", size=1) + 
    geom_histogram(binwidth = 3, col="gray5")  + 
    scale_fill_manual(values = myColorPalette) +
    # theme(legend.position="none") +
    ggtitle("STAI difference") +
    xlab("(T2 - T1)") +
facet_grid(rows=NULL, cols=vars(category))
```

#### STAI differences by site

```{r}
ggplot(data=staiDiff, aes(x=difference, fill=LOCATION)) +
    geom_vline(xintercept=0, color="gray5", size=1) + 
    geom_histogram(binwidth = 3, col="gray5")  + 
    scale_fill_manual(values = myColorPalette) +
    # theme(legend.position="none") +
    ggtitle("STAI difference") +
    xlab("(T2 - T1)") +
    facet_grid(rows=vars(LOCATION), cols=vars(category))
```


```{r}
sessionInfo()
```
