---
title: "Diversity"
output: html_document
date: "2023-09-15"
---

```{r setup, include=FALSE}
# I prefer to edit the Rmd file and then use purl to convert into a R script.
# knitr::purl(input=dir(pattern="Diversity_Metrics.Rmd"), documentation = 2L)
knitr::opts_chunk$set(echo = TRUE)
```

# ANIGMA cohort Taxa

### libraries

```{r}
suppressPackageStartupMessages({
  library(vegan)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})
```

A named vector of color vectors.  Most of these match the previous paper. 
```{r, fig.height=4, fig.width=2, echo=FALSE, include=FALSE}
themeFile = "../../input/themes.R"
if (file.exists(themeFile)){
    source(themeFile)
    myColorPalette = getColorPalette()
    displayColorPalette(myColorPalette)
}else{
    myColorPalette = c(
        # timepoint
        nonED = "#1B9E77",
        HC = "#1B9E77",
        T1 = "#D95F02",
        AN.AD = "#D95F02",
        T2 = "#7570B3",
        AN.DS = "#7570B3",
        # site
        Denver = "#1F78B4",
        ACUTE = "#1F78B4",
        UNC = "#F0027F",
        CEED = "#F0027F",
        FARGO = "coral",
        # subgroups indicating site AND group/timepoint
        UNC_HC = "#B2DF8A",
        Denver_HC = "#33A02C",
        UNC_T1 = "#FDBF6F",
        UNC_T2 = "#FF7F00",
        Denver_T1 = "#CAB2D6",
        Denver_T2 = "#6A3D9A"
    )
    # plot.new(); legend(legend=names(myColorPalette), 
    #                    col=myColorPalette, bty="n", x="center", pch=15, cex=2)
    
    colorKey = data.frame(color=myColorPalette, key=names(myColorPalette))
    # ggplot(data=colorKey, aes(x=1, y = 1:nrow(colorKey), fill=key, label=key)) +
    #     geom_tile() +
    #     scale_fill_manual(values = myColorPalette) +
    #     theme_void()+
    #     theme(legend.position="none") + 
    #     geom_text()
}

# Use the theme_classic() ggplot theme unless otherwise indicated.
theme_set(theme_classic())
```

Source the plot functions that go with this module.
```{r}
source("../resources/PlotDiversityRichness.R")
```


Pick a taxonomic level. Take an argument if one is given.
```{r}
args = commandArgs(trailingOnly=TRUE)
if (length(args) > 0){
    taxaLevel = args[1]
}else{
    taxaLevel = "species"
}
```

Direct output
```{r}
outDir = file.path("..", "output", taxaLevel)
suppressWarnings( dir.create(outDir, recursive = T) )
```

### Read counts data

```{r}
input=c(
    domain = "../../input/counts-tables/filtered-counts_level-1_domain.csv",
    phylum = "../../input/counts-tables/filtered-counts_level-2_phylum.csv",
    class = "../../input/counts-tables/filtered-counts_level-3_class.csv",
    order = "../../input/counts-tables/filtered-counts_level-4_order.csv",
    family = "../../input/counts-tables/filtered-counts_level-5_family.csv",
    genus = "../../input/counts-tables/filtered-counts_level-6_genus.csv",
    species = "../../input/counts-tables/filtered-counts_level-7_species.csv")

# taxaLevel = "class"
countsFile = input[taxaLevel]
```

We are looking at the ``r taxaLevel`` level, so we will read file ``r countsFile``.

```{r}
countsAndMetaALL = read.csv(countsFile)
dim(countsAndMetaALL)
```

Make a table with all the metadata that came with the artifact.
```{r}
qzaMeta = countsAndMetaALL %>% select(id.orig, StudyID, Timepoint)
```

Split the meta data and counts
```{r}
metaCols = c("id.orig", "BarcodeNumber", "BarcodeSequenceFull", "BarcodeSequence", "LinkerPrimerSequence", "StudyID", "Timepoint", "Person")
counts = countsAndMetaALL %>% select(-all_of(metaCols))
row.names(counts) = countsAndMetaALL$id.orig
key = countsAndMetaALL %>% select(id.orig, StudyID, Timepoint)
```

### Read meta data.

Read the original metadata file.
```{r}
meta = read.delim("../../input/meta/ANIGMA-metadata.txt") %>%
  select(PARTICIPANT.ID, LOCATION, TIMEPOINT, AGE, SUBTYPE, BMI, 
         STAI_Y1, STAI_Y2, STAI_TOTAL, PSS, DAYS_TREAT, Weight_kg, DNA.ID, DUR_ILLNESS_YRS) %>%
  filter(DNA.ID %in% countsAndMetaALL$id.orig)
row.names(meta) = meta$DNA.ID
dim(meta)
```

Check that the meta data that came with the qiime artifact matches up to the main project meta data.
```{r}
m = merge(meta, key, by.x="DNA.ID", by.y="id.orig")
table(m$PARTICIPANT.ID == m$StudyID)
table(m$TIMEPOINT == m$Timepoint)
```

The metadata from the qiime artifact and the metadata for the project match up. There is one that has an error in the qiime artifact metadata. That's fine. We'll use the project metadata moving forward.
```{r}
m[which(m$TIMEPOINT != m$Timepoint), c("PARTICIPANT.ID", "TIMEPOINT", "BMI", "Timepoint")]
```
## Merge

Merge the meta data with the counts data.
```{r}
m = merge(meta, key, by.x="DNA.ID", by.y="id.orig")
table(m$PARTICIPANT.ID == m$StudyID)
table(m$TIMEPOINT == m$Timepoint)
```

# MAIN

## Diversity 

Diversity on raw counts.
```{r}
div.alpha = vegan::diversity(counts, index="shannon")
meta$alpha.div.raw = div.alpha[row.names(meta)]
```

```{r plotDiv}
# source("../resources/PlotDiversityRichness.R")
# set defaults for this document
plotDiv = function(df, column, sig=0.05, saveToDir=outDir, tLevel=taxaLevel){
  plotDiversity(df, column=column, sig=sig, saveToDir=saveToDir, tLevel=tLevel)
}

plotDiv(meta, column="alpha.div.raw", sig=0.05)
```

Looking at the family level, HC is not very different from T1.  But T2 is less diverse.

### Split Diversity by site

The change in diversity is site specific.  Is that driven by the controls from each site or the AN from each site?

To test this, in each comparison, use all controls.

Save a pdf with these figures with matched next to each other:
```{r}
pdf(file=file.path(outDir, "diversity-by-site_joined-controls.pdf"))
plotDiv(meta %>% filter(LOCATION=="ACUTE"), column="alpha.div.raw", saveToDir=NA) + 
    ggtitle("ACUTE")
plotDiv(meta %>% filter(LOCATION=="ACUTE" | TIMEPOINT=="HC"), column="alpha.div.raw", saveToDir=NA) + 
    ggtitle("ACUTE - with all controls")

plotDiv(meta %>% filter(LOCATION=="CEED"), column="alpha.div.raw", saveToDir=NA) + 
    ggtitle("CEED")
plotDiv(meta %>% filter(LOCATION=="CEED" | TIMEPOINT=="HC"), column="alpha.div.raw", saveToDir=NA) + 
    ggtitle("CEED - with all controls")

plotDiv(meta %>% filter(LOCATION=="FARGO"), column="alpha.div.raw", saveToDir=NA) + 
    ggtitle("FARGO")
plotDiv(meta %>% filter(LOCATION=="FARGO" | TIMEPOINT=="HC"), column="alpha.div.raw", saveToDir=NA) + 
    ggtitle("FARGO - with all controls")
dev.off()
```


### Diversity on Rarefied counts

Diversity on rarified data.
```{r}
set.seed(95)

rareTo = min(rowSums(counts))
message("Rarefying each sample to depth: ", rareTo)

counts_rarified = vegan::rrarefy(counts, sample=rep(rareTo, nrow(counts)))

div.alpha.rare = vegan::diversity(counts_rarified, index="shannon")
meta$alpha.div.rarified = div.alpha.rare[row.names(meta)]
plotDiv(meta, column="alpha.div.rarified")
```


## Richness

richness on rarefied counts:
```{r}
countRichnessSimple <- function(sampleCounts){
  return(suppressWarnings(sum(sampleCounts > 0)))
}
rich.rarefied = apply(counts_rarified, countRichnessSimple, MARGIN=1)
meta$richness.rarefied = rich.rarefied[row.names(meta)]
```

```{r plotRich}
# source("../resources/PlotDiversityRichness.R")
# set defaults for this document
plotRich = function(df, column, sig=0.05, saveToDir=outDir, tLevel=taxaLevel){
    plotRichness(df, column, sig=sig, saveToDir=saveToDir, tLevel=tLevel, rareTo=rareTo)
}

plotRich(meta, column="richness.rarefied")
```


### Split Richness by site

Save a pdf of these figures with matched site next to each other.
```{r}
pdf(file=file.path(outDir, "richness-by-site_joined-controls.pdf"))

plotRich(meta %>% filter(LOCATION=="ACUTE"), column="richness.rarefied", saveToDir=NA) + 
    ggtitle("ACUTE")
plotRich(meta %>% filter(LOCATION=="ACUTE" | TIMEPOINT=="HC"), column="richness.rarefied", saveToDir=NA) + 
    ggtitle("ACUTE - with all controls")

plotRich(meta %>% filter(LOCATION=="CEED"), column="richness.rarefied", saveToDir=NA) + 
    ggtitle("CEED")
plotRich(meta %>% filter(LOCATION=="CEED" | TIMEPOINT=="HC"), column="richness.rarefied", saveToDir=NA) + 
    ggtitle("CEED - with all controls")

plotRich(meta %>% filter(LOCATION=="FARGO"), column="richness.rarefied", saveToDir=NA) + 
    ggtitle("FARGO")
plotRich(meta %>% filter(LOCATION=="FARGO" | TIMEPOINT=="HC"), column="richness.rarefied", saveToDir=NA) + 
    ggtitle("FARGO - with all controls")

dev.off()
```

# Save Results

```{r}
dataToSave = data.frame(DNA.ID = meta$DNA.ID,
                        shannon.diversity = meta$alpha.div.raw,
                        richness = meta$richness.rarefied)

filename = file.path(outDir, paste0("diversity-and-richness_", taxaLevel, ".txt"))
write.table(x=dataToSave, file=filename, sep="\t", quote=F, row.names = F)
message("Saved file to: ", filename)
```

```{r}
sessionInfo()
sizeByP
plotRichness
plotDiversity
```

