---
title: "Reshape Metadata"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Understand which values within the metadata are correlated. Make sure you understand how these things are related in reality and mathematically.

## Settup

### Libraries

```{r }
tellme <- function(name){message("Package ", name, " version: ", packageVersion(name))}

library(tidyr); tellme("tidyr")
suppressPackageStartupMessages(library(dplyr)); tellme("dplyr")
library(ggplot2); tellme("ggplot2")
```

Direct output.
```{r echo=FALSE}
outdir = "../output"
suppressWarnings( dir.create(outdir) )
```
Output will be saved to: ``r outdir``


```{r, fig.height=4, fig.width=2, echo=FALSE, include=FALSE}

### Colors - not currenlty used here

# A named vector of color vectors.  Most of these match the previous paper. 

themeFile = "../../input/themes.R"
if (file.exists(themeFile)){
    source(themeFile)
    myColorPalette = getColorPalette()
    displayColorPalette(myColorPalette)
}else{
    myColorPalette = c(
        # timepoint
        nonED = "#1B9E77",
        HC = "#1B9E77",
        T1 = "#D95F02",
        AN.AD = "#D95F02",
        T2 = "#7570B3",
        AN.DS = "#7570B3",
        # site
        Denver = "#1F78B4",
        ACUTE = "#1F78B4",
        UNC = "#F0027F",
        CEED = "#F0027F",
        FARGO = "coral",
        # subgroups indicating site AND group/timepoint
        UNC_HC = "#B2DF8A",
        Denver_HC = "#33A02C",
        UNC_T1 = "#FDBF6F",
        UNC_T2 = "#FF7F00",
        Denver_T1 = "#CAB2D6",
        Denver_T2 = "#6A3D9A"
    )
    # plot.new(); legend(legend=names(myColorPalette), 
    #                    col=myColorPalette, bty="n", x="center", pch=15, cex=2)
    
    colorKey = data.frame(color=myColorPalette, key=names(myColorPalette))
    ggplot(data=colorKey, aes(x=1, y = 1:nrow(colorKey), fill=key, label=key)) +
        geom_tile() +
        scale_fill_manual(values = myColorPalette) +
        theme_void()+
        theme(legend.position="none") + 
        geom_text()
}
```

### Read metadata

Get the per-particpant metadata.
```{r}
pipeline = "../../"
prevModule = dir(path=pipeline, pattern="Participant_Metadata", full.names = TRUE)
fileName = file.path(prevModule, "output", "ANIGMA-metadata_by_AN_participant.txt")
diffs = read.delim(fileName)
```

Read in data from file: ``r fileName``

This table has ``r nrow(diffs)`` rows and ``r ncol(diffs)`` columns.

Read per-sample data.
```{r}
meta = read.delim("../../input/meta/ANIGMA-metadata.txt") %>%
  select(PARTICIPANT.ID, LOCATION, TIMEPOINT, AGE, SUBTYPE, BMI, 
         STAI_Y1, STAI_Y2, STAI_TOTAL, PSS, DAYS_TREAT, Weight_kg, 
         DUR_ILLNESS_YRS)
dim(meta)
```

This meta data has ``r nrow(meta)`` rows and ``r ncol(meta)`` columns.

The HC equivalent data frame:
```{r}
HC = meta %>% filter(TIMEPOINT=="HC") %>%
  select(PARTICIPANT.ID, LOCATION, AGE, SUBTYPE, STAI_Y1, STAI_Y2, STAI_TOTAL, PSS, BMI, Weight_kg)
```

This subset of the data has ``r nrow(HC)`` rows and ``r ncol(HC)`` columns.

## Visualize broad view correlations

Create a correlation matrix.
```{r}
df = diffs %>% select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE)
cor.co = cor(df, use="pairwise.complete.obs", method = "pearson")
```

```{r}
# much of this is taken from
# http://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization

#functions
tidyMelt <- function(cormat, values_name="values"){
    cormat.new = cormat %>% data.frame() 
    cormat.new$var1 = row.names(cormat.new)
    melted_cormat = cormat.new %>% pivot_longer(cols=-var1, names_to = "var2", values_to = values_name)
    return(melted_cormat)
}

make_heatmap_1 <- function(cor.co2){
    ggplot(data = cor.co2, aes(x=var1, y=var2, fill=cor)) + 
        geom_tile(color = "white")+
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                             midpoint = 0, limit = c(-1,1), space = "Lab", 
                             name="Pearson\nCorrelation",
                             na.value = gray(.9)) +
        theme_minimal()+ # minimal theme
        xlab("") +
        ylab("") +
        theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                         size = 8, hjust = 1)) +
        coord_fixed()
}

cor.co2 = tidyMelt(cor.co, values_name="cor")

make_heatmap_1(cor.co2)
```

Use only the upper triangle, and order features in a sensible way.
```{r echo=FALSE}
# functions
# Get lower triangle of the correlation matrix
get_lower_tri<-function(cormat){
    cormat[upper.tri(cormat)] <- NA
    return(cormat)
}
# Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
}

reorder_cormat <- function(cormat){
    # Use correlation between variables as distance
    dd <- as.dist((1-cormat)/2)
    hc <- hclust(dd)
    cormat <-cormat[hc$order, hc$order]
}
```

```{r}
# actions

# Reorder the correlation matrix
cormat <- reorder_cormat(cor.co)
# upper_tri <- get_upper_tri(cormat)
triangle1 = get_upper_tri(cormat)
# Melt the correlation matrix
melted_cormat = tidyMelt(triangle1, values_name="cor")
# order levels to match triangle
melted_cormat$var1 = factor(melted_cormat$var1, levels=row.names(triangle1))
melted_cormat$var2 = factor(melted_cormat$var2, levels=row.names(triangle1))

# Create a ggheatmap
make_heatmap_1(melted_cormat)
```

Manually set the order.
```{r include=FALSE, echo=FALSE}
# get string from this:
# paste(row.names(triangle1), collapse = "', '")
# surround results like this: manual.order = c('<result>')
# then manually edit
```

```{r}
manual.order = c('BMI.T1', 'Weight_kg.T1', 'BMI.T2', 'Weight_kg.T2',
                 'T1.severity', 'DAYS_TREAT', 'BMI.diff', 'Weight_kg.diff', 
                 'AGE', 'DUR_ILLNESS_YRS', 
                 'PSS.diff', 'STAI_Y1.diff', 'STAI_Y2.diff', 'STAI_TOTAL.diff', 
                 'PSS.T2', 
                 'STAI_Y2.T1', 'PSS.T1', 'STAI_Y1.T1', 'STAI_TOTAL.T1', 'STAI_Y1.T2', 'STAI_Y2.T2', 'STAI_TOTAL.T2')
manual.order
```

Replot with manual order.
```{r}
# Reorder the correlation matrix
cormat <- cor.co[manual.order, manual.order]

# upper_tri <- get_upper_tri(cormat)
triangle1 = get_upper_tri(cormat)
# Melt the correlation matrix
melted_cormat = tidyMelt(triangle1, values_name="cor")
# order levels to match manual order
melted_cormat$var1 = factor(melted_cormat$var1, levels=row.names(triangle1))
melted_cormat$var2 = factor(melted_cormat$var2, levels=row.names(triangle1))

# Create a ggheatmap
make_heatmap_1(melted_cormat)
```

Now lets get some p-values. 
```{r}
feats = colnames(df)

pvalCut = 0.01

cors = matrix(data=NA, nrow=length(feats), ncol=length(feats), dimnames = list(feats, feats))
pvals = cors
cors.sig = cors

for (i in feats){
    for (j in feats){
        if ( i != j ){
            res = cor.test(df[,i], df[,j], method="pearson")
            cors[i,j] = res$estimate
            pvals[i,j] = res$p.value
            if (res$p.value < pvalCut){
                cors.sig[i,j] = res$estimate
            }
        }
    }
}
```

Make a new plot that only includes values were the correlation was significant (p < ``r pvalCut``).

```{r}
reshapeMatrixToPlot <- function(correlationMatrix){
    # Reorder the correlation matrix
    cormat <- correlationMatrix[manual.order, manual.order]
    
    # only one triangle
    triangle1 = get_upper_tri(cormat)
    
    # Melt the correlation matrix
    melted_cormat = tidyMelt(triangle1, values_name="cor")
    
    # order levels to match triangle order
    melted_cormat$var1 = factor(melted_cormat$var1, levels=row.names(triangle1))
    melted_cormat$var2 = factor(melted_cormat$var2, levels=row.names(triangle1))
    
    return(melted_cormat)
}

# match most recent plot - show melt and plot method matches
make_heatmap_1(reshapeMatrixToPlot(cor.co)) + ggtitle("Upper triangle")
# should also match most recent plot - show input matrix matches
make_heatmap_1(reshapeMatrixToPlot(cors)) + ggtitle("Upper triangle, excluding identity line")
# only sig
make_heatmap_1(reshapeMatrixToPlot(cors.sig)) + ggtitle(paste0("Correlations with p-value less than ", pvalCut))
```

Now lets note what is mathmatically related.
```{r}
# T1 severity is related to BMI at T1 and therefore also related to weight at T1
severity = data.frame(x=c("T1.severity", "T1.severity"), 
                      y=c("BMI.T1", "Weight_kg.T1"))

related = data.frame(x=c("STAI_TOTAL", "STAI_TOTAL", "BMI"),
                     y=c("STAI_Y1", "STAI_Y2", "Weight_kg"))
# these are related within a given time point
related1 = data.frame(x=paste(related$x, "T1", sep="."),
                      y=paste(related$y, "T1", sep="."))
related2 = data.frame(x=paste(related$x, "T2", sep="."),
                      y=paste(related$y, "T2", sep="."))

# by extension, the diff of each thing is related to each time point of the other thing
related1.diff = data.frame(x=paste(related$x, "T1", sep="."),
                      y=paste(related$y, "diff", sep="."))
related1.diff.rev = data.frame(x=paste(related$x, "diff", sep="."),
                      y=paste(related$y, "T1", sep="."))
related2.diff = data.frame(x=paste(related$x, "T2", sep="."),
                      y=paste(related$y, "diff", sep="."))
related2.diff.rev = data.frame(x=paste(related$x, "diff", sep="."),
                      y=paste(related$y, "T2", sep="."))

# all diffables have a relationship between their diff and each time point
# diffables = c("STAI_Y1", "STAI_Y2", "STAI_TOTAL", "PSS", "BMI", "Weight_kg")
diffables = gsub(grep(".diff", names(diffs), value = T), pattern=".diff", replacement="")
relatedDiff1 = data.frame(x=paste(diffables, "T1", sep="."),
                         y=paste(diffables, "diff", sep="."))
relatedDiff2 = data.frame(x=paste(diffables, "T2", sep="."),
                         y=paste(diffables, "diff", sep="."))

stack = rbind(related1, related2, 
              related1.diff, related1.diff.rev, 
              related2.diff, related2.diff.rev, 
              relatedDiff1, relatedDiff2, 
              severity)

# relatedness goes both ways
reverse.stack = data.frame(x=stack$y, y=stack$x)
stack2 = rbind(stack, reverse.stack)

head(stack2)
```

test annotations
```{r}
make_heatmap_1(reshapeMatrixToPlot(cor.co)) + 
    # mark related features
    annotate(geom="text", x=stack2$x, y=stack2$y, label="+") +
    # identity line gets similar marking, slighly bigger
    annotate(geom="text", x=feats, y=feats, label="+", size=6)
```

### Finalize figure

Show all correlations on the upper triangle but in the lower triangle show only the significant ones.
```{r}
# upper - all correlation values
# Reorder the correlation matrix
cormat <- cor.co[manual.order, manual.order]
# only one triangle
triangle1 = get_upper_tri(cormat)
# Melt the correlation matrix
melted_cormat1 = tidyMelt(triangle1, values_name="cor") %>% filter(!is.na(cor))
# order levels to match triangle order
melted_cormat1$var1 = factor(melted_cormat1$var1, levels=manual.order)
melted_cormat1$var2 = factor(melted_cormat1$var2, levels=manual.order)

# lower - only sig values
cormat <- cors.sig[manual.order, manual.order]
# only one triangle
get_lower_tri2<-function(cormat){
    cormat[upper.tri(cormat)] <- 50
    for (f in feats){
        cormat[f, f] <- 50
    }
    return(cormat)
}
triangle2 = get_lower_tri2(cormat)

# Melt the correlation matrix
melted_cormat2 = tidyMelt(triangle2, values_name="cor") %>% filter(cor != 50)
# order levels to match triangle order
melted_cormat2$var1 = factor(melted_cormat2$var1, levels=manual.order)
melted_cormat2$var2 = factor(melted_cormat2$var2, levels=manual.order)

# merge in melted form
melted_cormat = rbind(melted_cormat1, melted_cormat2)
# order levels to match triangle order
melted_cormat$var1 = factor(melted_cormat$var1, levels=manual.order)
melted_cormat$var2 = factor(melted_cormat$var2, levels=manual.order)

figure = make_heatmap_1(melted_cormat) + 
    # mark related features
    annotate(geom="text", x=stack2$x, y=stack2$y, label="+") +
    # identity line gets similar marking, slighly bigger
    annotate(geom="text", x=feats, y=feats, label="+", size=6)
print(figure)
```


## Figure Variants

The objects we need moving forward are:

 * markPairings - a data frame with 2 columns indicating pairs of items to mark as mathmatically related.
 * my.order - a vector giving the order in which items should be plotted.
 * dataframe - a data frame with samples in rows and columns of features to compare for correalations.
 * pvalCut - a p-value threshold for the lower triangle
 
Remove everything else.
```{r}
rm(list = setdiff(ls(), c("manual.order", "stack2", "diffs", "HC", "outdir")))
```

 

As a function:
```{r}
# helper functions
# Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
}

tidyMelt <- function(cormat, values_name="values"){
    cormat.new = cormat %>% data.frame() 
    cormat.new$var1 = row.names(cormat.new)
    melted_cormat = cormat.new %>% pivot_longer(cols=-var1, names_to = "var2", values_to = values_name)
    return(melted_cormat)
}

make_heatmap_1 <- function(cor.co2){
    ggplot(data = cor.co2, aes(x=var1, y=var2, fill=cor)) + 
        geom_tile(color = "white")+
        scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                             midpoint = 0, limit = c(-1,1), space = "Lab", 
                             name="Pearson\nCorrelation",
                             na.value = gray(.9)) +
        theme_minimal()+ # minimal theme
        xlab("") +
        ylab("") +
        theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                         size = 8, hjust = 1)) +
        coord_fixed()
}
```

```{r}
# big function
make_heatmap_2 <- function(dataframe, my.order=manual.order, markPairings=stack2, pvalCut=0.01,
                           cor.mat.file=NULL, pval.mat.file=NULL){
    if (is.null(my.order)){
        my.order = names(dataframe)
    }
    
    feats = colnames(dataframe)
    
    if (!is.null(markPairings)){
        names(markPairings) = c("x", "y")
        markPairings = markPairings %>% 
            filter( x %in% feats) %>%
            filter( y %in% feats)
    }
    
    cor.co = cor(dataframe, use="pairwise.complete.obs", method = "pearson")
    if (!is.null(cor.mat.file)){
        write.table(cbind(feature=row.names(cor.co), cor.co), 
                    file=cor.mat.file, 
                    sep="\t", quote=F, row.names = F, col.names = T)
    }
    
    cors = matrix(data=NA, nrow=length(feats), 
                  ncol=length(feats), 
                  dimnames = list(feats, feats))
    pvals = cors
    cors.sig = cors
    
    for (i in feats){
        for (j in feats){
            if ( i != j ){
                res = cor.test(dataframe[,i], dataframe[,j], method="pearson")
                pvals[i,j] = res$p.value
                if (res$p.value < pvalCut){
                    cors.sig[i,j] = res$estimate
                }
            }
        }
    }
    
    if (!is.null(pval.mat.file)){
        write.table(cbind(feature=row.names(pvals), pvals), 
                    file=pval.mat.file, 
                    sep="\t", quote=F, row.names = F, col.names = T)
    }
    
    # upper - all correlation values
    # Reorder the correlation matrix
    cormat <- cor.co[my.order, my.order]
    # only one triangle
    triangle1 = get_upper_tri(cormat)
    # Melt the correlation matrix
    melted_cormat1 = tidyMelt(triangle1, values_name="cor") %>% filter(!is.na(cor))
    # order levels to match triangle order
    melted_cormat1$var1 = factor(melted_cormat1$var1, levels=my.order)
    melted_cormat1$var2 = factor(melted_cormat1$var2, levels=my.order)
    
    # lower - only sig values
    cormat <- cors.sig[my.order, my.order]
    # only one triangle
    get_lower_tri2<-function(cormat){
        cormat[upper.tri(cormat)] <- 50
        for (f in feats){
            cormat[f, f] <- 50
        }
        return(cormat)
    }
    triangle2 = get_lower_tri2(cormat)
    
    # Melt the correlation matrix
    melted_cormat2 = tidyMelt(triangle2, values_name="cor") %>% filter(cor != 50)
    # order levels to match triangle order
    melted_cormat2$var1 = factor(melted_cormat2$var1, levels=my.order)
    melted_cormat2$var2 = factor(melted_cormat2$var2, levels=my.order)
    
    # merge in melted form
    melted_cormat = rbind(melted_cormat1, melted_cormat2)
    # order levels to match triangle order
    melted_cormat$var1 = factor(melted_cormat$var1, levels=my.order)
    melted_cormat$var2 = factor(melted_cormat$var2, levels=my.order)
    
    figure = make_heatmap_1(melted_cormat)
    
    if (!is.null(markPairings)){
        figure = figure + 
            # mark related features
            annotate(geom="text", x=markPairings[,1], y=markPairings[,2], label="+") +
            # identity line gets similar marking, slighly bigger
            annotate(geom="text", x=feats, y=feats, label="+", size=6)
    }
    
    return(figure)
}

fig1 = make_heatmap_2(diffs %>% select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE),
                      cor.mat.file = file.path(outdir, "correlations_AN_cor.txt"),
                      pval.mat.file = file.path(outdir, "correlations_AN_pval.txt"))
fig1
```

Save figure to file.
```{r}
ggsave(filename = file.path(outdir, "correlations_AN.png"),
       plot = fig1)
```

Add LOCATION and SUBTYPE as a row.
```{r}
make_heatmap_2(diffs %>% 
                   mutate(LOC.N = as.numeric(factor(LOCATION))) %>%
                   mutate(TYPE.N = as.numeric(factor(SUBTYPE))) %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE),
               my.order = c( "TYPE.N", "LOC.N", manual.order))
```

### AN Subtypes

```{r}
s1 = make_heatmap_2(diffs %>%
                   filter(SUBTYPE == "BP") %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE)) +
    ggtitle("BP")

s2 = make_heatmap_2(diffs %>%
                   filter(SUBTYPE == "ANR") %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE)) +
    ggtitle("ANR")

## not enough of these to plot
# 
# make_heatmap_2(diffs %>%
#                    filter(SUBTYPE == "EDNOS") %>%
#                    select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE)) +
#     ggtitle("EDNOS")
# 
# make_heatmap_2(diffs %>%
#                    filter(SUBTYPE == "ARFID") %>%
#                    select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE)) +
#     ggtitle("ARFID")
```

```{r}
fileName.locations = file.path(outdir, "correlations_by-subtype.pdf")
pdf(fileName.locations)

s1
s2

dev.off()

s1
s2
```

### Location subsets

Using the above function.

```{r}
p1 = make_heatmap_2(diffs %>%
                   filter(LOCATION == "ACUTE") %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE)) +
    ggtitle("ACUTE")


# no ceed values for DUR_ILLNESS_YRS
p2 = make_heatmap_2(diffs %>%
                   filter(LOCATION == "CEED") %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE, -DUR_ILLNESS_YRS),
               my.order = setdiff(manual.order, "DUR_ILLNESS_YRS") ) +
    ggtitle("CEED")


diffs.FARGO = diffs %>%
    filter(LOCATION == "FARGO") %>%
    select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE) %>%
    select(-contains("STAI_Y2"), -contains("STAI_TOTAL") )
p3 = make_heatmap_2(diffs.FARGO,
               my.order=manual.order[manual.order %in% names(diffs.FARGO)],
               markPairings = stack2 %>% 
                   filter(x %in% names(diffs.FARGO)) %>%
                   filter(y %in% names(diffs.FARGO))) +
    ggtitle("FARGO")
```


```{r}
fileName.locations = file.path(outdir, "correlations_by-location.pdf")
pdf(fileName.locations)

p1
p2
p3

dev.off()

p1
p2
p3
```


### HC group

```{r}
hc1 = make_heatmap_2(HC %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE),
               my.order = NULL, markPairings = stack2) +
    ggtitle("Non-eating Disorder")
hc1
```


```{r}
hc2 = make_heatmap_2(HC %>%
                   filter(LOCATION == "ACUTE") %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE),
               my.order = NULL) +
    ggtitle("Non-eating Disorder - ACUTE")


hc3 = make_heatmap_2(HC %>%
                   filter(LOCATION == "CEED") %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE),
               my.order = NULL) +
    ggtitle("Non-eating Disorder - CEED")

hc4 = make_heatmap_2(HC %>%
                   filter(LOCATION == "FARGO") %>%
                   select(-PARTICIPANT.ID, -LOCATION, -SUBTYPE) %>%
                   select(-contains("STAI_Y2"), -contains("STAI_TOTAL")),
               my.order = NULL) +
    ggtitle("Non-eating Disorder - FARGO")

```

```{r}
fileName.locations = file.path(outdir, "correlations_non-ed_HC.pdf")
pdf(fileName.locations)

hc1
hc2
hc3
hc4

dev.off()

hc2
hc3
hc4
```


```{r}
sessionInfo()
```
