---
title: "Taxonomy vs Feature"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
---

```{r include=FALSE}
# I prefer to edit the Rmd file and then use purl to convert into a R script.
# knitr::purl(input=dir(pattern="Rmd"), documentation = 2L)
knitr::opts_chunk$set(echo = TRUE)
```

Core questions:

Does any one taxon have a significant correlation with any of the continuous metadata variables?
Does any one taxon have a significant difference between any two categories in the categorical metadata variables?

# Setup

### libraries

```{r results = "hold"}
tellme <- function(name){print(paste0("Package ", name, " version: ", packageVersion(name)))}

library(tidyr); tellme("tidyr")
suppressPackageStartupMessages(library(dplyr)); tellme("dplyr")
library(ggplot2); tellme("ggplot2")
library(ggrepel); tellme("ggrepel")
```
### Colors

A named vector of color vectors.  Most of these match the previous paper. 
```{r, fig.height=4, fig.width=2, echo=FALSE, include=FALSE}
themeFile = "../../input/themes.R"
if (file.exists(themeFile)){
    source(themeFile)
    myColorPalette = getColorPalette()
    displayColorPalette(myColorPalette)
}else{
    myColorPalette = c(
        # timepoint
        nonED = "#1B9E77",
        HC = "#1B9E77",
        T1 = "#D95F02",
        AN.AD = "#D95F02",
        T2 = "#7570B3",
        AN.DS = "#7570B3",
        # site
        Denver = "#1F78B4",
        ACUTE = "#1F78B4",
        UNC = "#F0027F",
        CEED = "#F0027F",
        FARGO = "coral",
        # subgroups indicating site AND group/timepoint
        UNC_HC = "#B2DF8A",
        Denver_HC = "#33A02C",
        UNC_T1 = "#FDBF6F",
        UNC_T2 = "#FF7F00",
        Denver_T1 = "#CAB2D6",
        Denver_T2 = "#6A3D9A"
    )
    # plot.new(); legend(legend=names(myColorPalette), 
    #                    col=myColorPalette, bty="n", x="center", pch=15, cex=2)
    
    colorKey = data.frame(color=myColorPalette, key=names(myColorPalette))
    # ggplot(data=colorKey, aes(x=1, y = 1:nrow(colorKey), fill=key, label=key)) +
    #     geom_tile() +
    #     scale_fill_manual(values = myColorPalette) +
    #     theme_void()+
    #     theme(legend.position="none") + 
    #     geom_text()
}

# Use the theme_classic() ggplot theme unless otherwise indicated.
theme_set(theme_classic())
```

Direct output
```{r}
outDir = file.path("..", "output")
suppressWarnings( dir.create(outDir, recursive = T) )
```

Output will be saved to: ``r outDir``

# Main

## Read meta data.

Read the original metadata file.
```{r}
meta = read.delim("../../input/meta/ANIGMA-metadata.txt") %>%
  select(PARTICIPANT.ID, LOCATION, TIMEPOINT, AGE, SUBTYPE, BMI, 
         STAI_Y1, STAI_Y2, STAI_TOTAL, PSS, DAYS_TREAT, Weight_kg, DNA.ID, DUR_ILLNESS_YRS) %>%
  mutate(linker = paste0("X", PARTICIPANT.ID, TIMEPOINT))
  # filter(DNA.ID != "") 
row.names(meta) = meta$linker
dim(meta)
```

Check metadata.
```{r}
if (any(!is.na(meta$DAYS_TREAT[meta$TIMEPOINT=="HC"]))){
    message("For any Healthy Control participant, the DAYS_TREAT should be NA (not 0).")
    meta$DAYS_TREAT[meta$TIMEPOINT=="HC"] = NA
}else{
    message("Ah, good, for Healthy Control participants, the DAYS_TREAT is NA (not 0).")
}
```

## Read pathway data

anywhere that references the 'taxaLevel' variable, say "pathway".
```{r}
taxaLevel="pathway"
```


```{r}
inputFile="../../input/pathway/humanN2_pathabundance_cpm.tsv"
```

Read file ``r inputFile``.

```{r}
pth = read.delim2(inputFile)
dim(pth)
```

```{r}
pth.info = select(pth, X..Pathway)
pth.info$pathway = sapply(strsplit(pth.info$X..Pathway, split="|", fixed=T), "[", 1)
pth.info$lineage = sapply(strsplit(pth.info$X..Pathway, split="|", fixed=T), "[", 2)
pth.info$path.id = sapply(strsplit(pth.info$pathway, split=":", fixed=T), "[", 1)
pth.info$name = sapply(strsplit(pth.info$pathway, split=": ", fixed=T), "[", 2)
# head(pth.info, 3)
```

TODO: Create a version of pth where the values are summed by pathway id.
```{r}
# pthSums = pth %>%
#   select(-X..Pathway) %>%
#   mutate_all(function(x) as.numeric(as.character(x))) %>%
#   aggregate(by=list(path.id = pth.info$path.id), FUN=sum)
# row.names(pthSums) = pthSums$path.id
# pthSums = select(pthSums, -path.id)
```

Extract the rows from the pathway info where there is no "|", so as to only get the pathway summary.
```{r}
pthSums = pth %>% 
  filter(X..Pathway != "UNMAPPED") %>% 
  filter(X..Pathway != "UNINTEGRATED") %>%
  filter(!grepl("|", X..Pathway, fixed = T)) 
row.names(pthSums) = pthSums$X..Pathway
pthSums = select(pthSums, -X..Pathway)
```


Extract ids from the column names that we can use to match up to the meta data.

```{r}
samples = data.frame(long = names(select(pth, -X..Pathway)) )
samples$short = sapply(strsplit(samples$long, split="_", fixed=T), "[", 1)
head(samples, 3)
```

Merge the counts and metadata.
```{r}
m = merge(samples, meta, by.x="short", by.y="linker")
head(m[c("long", "short", "DNA.ID")], 3)

newSampleNames = m$DNA.ID
names(newSampleNames) = m$long
```


Apply shorter sample names.
```{r}
# path.counts = pthSums %>% select(pthSums, all_of(m$long)) %>% t()
path.counts = t(pthSums[,m$long]) %>%
  as.data.frame() %>%
  mutate_all(as.numeric)
row.names(path.counts) = newSampleNames # order matched to m$long
write.table(file=file.path(outDir, "path.counts.tsv"), x=cbind(sample=row.names(path.counts), path.counts), row.names = F, col.names = T, quote=F, sep="\t")
path.counts[1:3, 1:3]
# 
# path.counts2 = apply(path.counts, 1, as.numeric)
# names(path.counts2) = names(path.counts)
# row.names(path.counts2) = names(path.counts)
```

TODO: move this code to line up names to its own script and just read the nice file here to keep this script simple.

Now we have a table of pathway counts in the same format that we had for a taxonomy table.

Which samples were lost because they do not have a matching row in the meta data?
```{r}
samples$short[which(! samples$short %in% m$short)]
```

### Filtering

Some of these pathways only have very little data, sometimes none. 

Remove the ones that are only detected at all in a few samples.
```{r}
samplesPerPathway = apply(path.counts, 2, function(x) sum(x>0))
path.counts = path.counts[,samplesPerPathway > 50]
dim(path.counts)
```




### normalize counts

Normalize counts.
```{r}
lognorm <- function(table, log10.do = TRUE){
  # table - a table with rows for samples and columns for features
  #         samples names are row names.
  #         feature names are column names.
  #         the table values are all numeric
  sampleSums = rowSums(table)
  meanSampleDepth = mean(sampleSums)
  sampleProportion = t( apply(table, 1, function(row) row / sum(row)) )
  t2 = sampleProportion * meanSampleDepth
  if (log10.do) t3 = log10( t2 + 1 )
  else t3 = t2 + 1
  return( t3 )
}

# counts.norm = lognorm(counts)

counts.norm = lognorm(path.counts)
```






### Feature engineering

For subtype, we are only interested in the AN cases. HC is not really an AN subtype. And we need to split samples between T1 and T2 so we are not double-dipping with each paired sample.
```{r}
meta$SUBTYPE.AN = meta$SUBTYPE
meta$SUBTYPE.AN[meta$SUBTYPE.AN=="HC"] = NA
```

The anorexia signal is what we would see by comparing T1 to HC or T2 to HC, which is the "TIMEPOINT" variable. Lets call it the ANOREXIA variable so that its clearer in the summaries.
```{r}
meta$ANOREXIA = meta$TIMEPOINT
```


### Previous feature engineering

Read the diffs metadata - data where diffs have been calculated.
```{r}
prevModule = dir(path="../..", pattern="Participant_Metadata", full.names = T)
diffsFile = file.path(prevModule, "output", "ANIGMA-metadata_by_AN_participant.txt")
diffs = read.delim(diffsFile)
row.names(diffs) = diffs$PARTICIPANT.ID
dim(diffs)
```

Read in modified metadata from: ``r diffsFile``

This table has ``r nrow(diffs)`` rows and ``r ncol(diffs)`` columns.

Match this data withe corresponding DNA ID for T1 and T2.
```{r}
diffKey = meta %>% 
    select(PARTICIPANT.ID, TIMEPOINT, DNA.ID) %>%
    pivot_wider(id_cols = PARTICIPANT.ID, names_from = TIMEPOINT, values_from = DNA.ID, names_prefix = "DNA.ID.") %>%
    select(-DNA.ID.HC)
diffs = merge(diffs, diffKey, all.x=T, by="PARTICIPANT.ID")
```


## merge

Merge the metadata and the normalized counts data.
```{r}
data = merge(counts.norm, meta, by.x=0, by.y="DNA.ID")
dim(data)
```

Make a scrambled version. We can use this to verify that we see null results, and our tests are not inherently flawed.
```{r}
scram.counts = counts.norm
row.names(scram.counts) = sample(row.names(scram.counts), size=nrow(scram.counts), replace = FALSE)
scrambled.data = merge(scram.counts, meta, by.x=0, by.y="DNA.ID")
rm(scram.counts)
```

Merge the T1 counts data to the per-participant differences data.
```{r}
diffData1 = merge(counts.norm, diffs, by.x=0, by.y="DNA.ID.T1")
diffData2 = merge(counts.norm, diffs, by.x=0, by.y="DNA.ID.T2")
```


Make sure Patient ID is not treated as a numerical value.
```{r}
data$PARTICIPANT.ID = as.character(data$PARTICIPANT.ID)
# scrambled.data$PARTICIPANT.ID = as.character(scrambled.data$PARTICIPANT.ID)
```

# Test

## test types

Group the metadata variables based on how they should be tested.

We have our categorical variables (two or more categories). These can be split in to variables that should be handled across time (T1 and T2 samples are both included in the same test) or by time (the test should be exclude one or the other time point so all samples in the test are independent). 

T1 and T2, but not HC.
```{r}
# variables_acrossTime = c("PARTICIPANT.ID", "TIMEPOINT")
variables_acrossTime = c("TIMEPOINT")
variables_acrossTime
```
Handle by time: T1 (HC and T1) or T2 (HC and T2)
```{r}
variables_byTime = c("LOCATION", "ANOREXIA", "SUBTYPE.AN")
variables_byTime
```


A subset of those are binary, and we can use the t-test.
```{r}
binary = c("TIMEPOINT.AN")
# HC vs T1
# HC vs T2
binary
```

Continuous variables come in two forms: 

 * constants - measurements that are assumed to be the same for a given person at T1 and T2, such as age.
 * diffables - measurements where the difference could be calculated and is potentially meaningful. These are metrics that:
 
   - could be tested at either time point, and 
   - should not be tested with both time points simultaneously because there are many pairs of non-independent samples, and 
   - we could test difference against taxa in addition to testing the T1 and the T2 values.
 
```{r}
constants = c("AGE", "DUR_ILLNESS_YRS", "DAYS_TREAT")
```
 
```{r}
diffables =c("Weight_kg", "BMI", "STAI_Y1", "STAI_Y2", "STAI_TOTAL", "PSS")
diffables
```

We have previously calculated differences, and a couple other per-patient numerical features: T1.severity and BMI.gain.per.day.
```{r}
# deltas = names(diffs) %>% grep(pattern="diff", value = T)
deltas = c("T1.severity", "Weight_kg.diff", "BMI.diff", "BMI.gain.per.day", "STAI_Y1.diff", "STAI_Y2.diff", "STAI_TOTAL.diff", "PSS.diff")
deltas
```

taxa
```{r}
taxa = colnames(counts.norm)
```

We have ``r length(taxa)`` to test.

## Categorical

Compare the values for each taxon against each categorical feature. Use the 1-way-anova and the Kruskal test.

```{r TaxaVsCategoryTest}
displayTaxonName <- function(string, string2){
  return(string)
}


TaxaVsCategoryTest <- function(dataDF, taxa, variables, doPlot=TRUE, redPval=0.05, testType="anova", fileNameBase=NULL, minGroupSize=2){
    # testType - one of "anova" or "kruskal.test"
    # taxa - a subset of columns representing taxonomic features
    # variables - a different subset of columns
    
    # check args
    if (length(setdiff(taxa, names(dataDF))) > 0) {
        stop("Argument 'taxa' should be a subset of the column names of 'dataDF'.")
    }
    if (length(setdiff(variables, names(dataDF))) > 0) {
        stop("Argument 'variables' should be a subset of the column names of 'dataDF'.")
    }
    
    pvalTable = data.frame(matrix(data=NA, nrow=length(taxa), ncol=length(variables)))
    names(pvalTable) = variables
    row.names(pvalTable) = taxa
    
    # adjusted pvalues have the same layout
    adjPvalTable = pvalTable
    
    message("p-values will be calculated based on the ", testType, " test.")
    saveToFile = !is.null(fileNameBase)
    
    pvalCountTable = data.frame(variable="TOTAL",
                                testType=testType,
                                nVals=NA,
                                levels=NA,
                                testWithPvalue=NA,
                                rawPvalSig=NA)
    
    for (variable in variables){
        variablePlotList = list()
        variablePvals = c()
        
        xAxis = dataDF[, variable]
        
        # if any category only has one sample (or fewer than minGroupSize), omit that category.
        tooFew = names(which(table(xAxis) < minGroupSize))
        if (length(tooFew) > 0){
            xAxis[xAxis %in% tooFew] = NA
            message("Omitted groups with fewer than [", minGroupSize, "] samples: ", paste0(tooFew, collapse=", "))
        }
        
        for (taxon in taxa){
            taxonVals = dataDF[,taxon]
            # message("Testing variable ", variable, " against taxon ", taxon, "...")
            
            if (testType == "anova"){
                pval = anova(lm(taxonVals ~ xAxis))$`Pr(>F)`[1]
            }
            if (testType == "kruskal.test"){
                pval = kruskal.test( taxonVals ~ xAxis, na.action=na.exclude )$p.value
            }
            
            pvalTable[taxon, variable] = pval
            variablePvals[taxon] = pval
            
            if (doPlot){
                plotdf = data.frame(x=xAxis, y=dataDF[,taxon], TIMEPOINT=dataDF$TIMEPOINT)
                gplot = ggplot(data=plotdf, aes(x=x, y=y, col=x)) +
                    xlab(variable) +
                    ylab(displayTaxonName(taxon, taxaLevel)) +
                    geom_boxplot(outlier.alpha = 0) +
                    # geom_point() +
                    geom_jitter(position=position_jitterdodge(jitter.width = .09, jitter.height = 0), size=.5) + 
                    scale_colour_manual(variable, values = myColorPalette) +
                    ggtitle( paste0(testType, " p-value: ", signif(pval, 5)) ) +
                    theme(plot.title = element_text(colour = ifelse(pval<=redPval, "red", "black"))) 
                variablePlotList[[taxon]] = gplot
            }
        }# taxon for loop
        
        
        # adjustedPvals
        padj.method="fdr"
        adjustedPvals = p.adjust(variablePvals, method = padj.method)
        adjPvalTable[,variable] = adjustedPvals
        
        # add row to summary table
        pvalCountTable = rbind(pvalCountTable, 
                               c(
                                   variable=variable,
                                   testType=testType,
                                   nVals=sum(!is.na(xAxis)),
                                   levels=paste0(unique(xAxis), collapse=","),
                                   testWithPvalue=sum(!is.na(variablePvals)),
                                   rawPvalSig=sum(variablePvals <= redPval, na.rm=T)
                               ) 
        )
        
        
        if (doPlot){
            ord = order(variablePvals)
            if(saveToFile) {
                pdfname = paste0(fileNameBase, "_plots_", variable, ".pdf")
                pdf(pdfname)
                message("Saving file: ", pdfname)
            }
            for (i in ord){
                annotColor = ifelse(adjustedPvals[i] <= redPval, "red", "black")
                annotText = paste0(padj.method, "-adjusted p-value: ", signif(adjustedPvals[i], 5))
                show(
                    variablePlotList[[i]] +
                        annotate(geom="text", label=annotText, col=annotColor,
                                 x=Inf, y=Inf, hjust="inward", vjust="inward")
                )
            }
            if(saveToFile) {
                dev.off()
            }
        }
        
    }# variable for-loop
    
    if (saveToFile){
        # save summary table
        table.file = paste0(fileNameBase, "_summary.txt")
        message("Saving file: ", table.file)
        write.table(file=table.file, pvalCountTable, sep="\t", quote=F, row.names = F)
    }
    
    return(pvalTable)
}
```

```{r runTaxaVsCategoryTest}
# parametric test - anova
acrossTimeTable = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "HC"),
                                     taxa, variables_acrossTime,
                                     fileNameBase=file.path(outDir, "T1T2.anova"))
acrossTimeTable.acute = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "HC" & LOCATION=="ACUTE"),
                                     taxa, variables_acrossTime,
                                     fileNameBase=file.path(outDir, "T1T2.ACUTE.anova"))
acrossTimeTable.ceed = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "HC" & LOCATION=="CEED"),
                                     taxa, variables_acrossTime,
                                     fileNameBase=file.path(outDir, "T1T2.CEED.anova"))
acrossTimeTable.fargo = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "HC" & LOCATION=="CEED"),
                                     taxa, variables_acrossTime,
                                     fileNameBase=file.path(outDir, "T1T2.FARGO.anova"))

# randomized.acrossTimeTable = TaxaVsCategoryTest(scrambled.data, taxa, variables_acrossTime, doPlot=F)

anovaAtT1 = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "T2"), 
                               taxa, variables_byTime, minGroupSize = 5,
                               fileNameBase=file.path(outDir, "HCT1.anova"))

anovaAtT2 = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "T1"), 
                               taxa, variables_byTime, minGroupSize = 5, 
                               fileNameBase=file.path(outDir, "HCT2.anova"))

anovaAtHC = TaxaVsCategoryTest(data %>% filter(TIMEPOINT == "HC"), 
                               taxa, "LOCATION", 
                               fileNameBase=file.path(outDir, "HC.anova"))

# nonparametric test
kwAcrossTime = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "HC"),
                                  taxa, variables_acrossTime, doPlot=F, testType = "kruskal.test")
# randomized.kwAcrossTime = TaxaVsCategoryTest(scrambled.data, taxa, variables_acrossTime, doPlot=F, testType = "kruskal.test")
kwAtT1 = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "T2"), 
                            taxa, variables_byTime, doPlot=F, testType = "kruskal.test")
kwAtT2 = TaxaVsCategoryTest(data %>% filter(TIMEPOINT != "T1"), 
                            taxa, variables_byTime, doPlot=F, testType = "kruskal.test")
kwAtHC = TaxaVsCategoryTest(data %>% filter(TIMEPOINT == "HC"), 
                            taxa, "LOCATION", doPlot=F, testType = "kruskal.test")
```


## binary tests - t-test

This is basically the same as the anova test with two categories, so we my not bother.

## Numeric feature tests

Compare the values for each taxon against each numeric feature.

```{r TaxaVsNumericTest}
TaxaVsNumericTest <- function(dataDF, taxa, variables, doPlot=TRUE, redPval=0.05, testType="pearson", fileNameBase=NULL){
    # testType - one of "pearson" or "kendall"
    # taxa - a subset of columns representing taxonomic features
    # variables - a different subset of columns
    
    # check args
    if (length(setdiff(taxa, names(dataDF))) > 0) {
        stop("Argument 'taxa' should be a subset of the column names of 'dataDF'.")
    }
    if (length(setdiff(variables, names(dataDF))) > 0) {
        stop("Argument 'variables' should be a subset of the column names of 'dataDF'.")
    }
    
    pvalTable = data.frame(matrix(data=NA, nrow=length(taxa), ncol=length(variables)))
    names(pvalTable) = variables
    row.names(pvalTable) = taxa
    
    # adjusted pvalues have the same layout
    adjPvalTable = pvalTable
    
    message("p-values will be calculated based on the ", testType, " test.")
    saveToFile = !is.null(fileNameBase)
    
    pvalCountTable = data.frame(variable="TOTAL",
                                testType=testType,
                                nVals=NA,
                                levels=NA,
                                testWithPvalue=NA,
                                rawPvalSig=NA)
    
    for (variable in variables){
        variablePlotList = list()
        variablePvals = c()
        
        xAxis = dataDF[, variable]
        
        for (taxon in taxa){
            taxonVals = dataDF[,taxon]
            # message("Testing variable ", variable, " against taxon ", taxon, "...")
            xAxis = dataDF[, variable]
            if (testType == "pearson"){
                pval = cor.test( taxonVals, xAxis, na.action=na.exclude, method="pearson")$p.value
            }
            if (testType == "kendall"){
                pval = cor.test( taxonVals, xAxis, na.action=na.exclude, method="kendall")$p.value
            }
            
            pvalTable[taxon, variable] = pval
            variablePvals[taxon] = pval
            
            if (doPlot){
                plotdf = data.frame(x=dataDF[,variable], y=dataDF[,taxon], LOCATION=dataDF$LOCATION)
                gplot = ggplot(data=plotdf, aes(x=x, y=y, col=LOCATION)) +
                    xlab(variable) +
                    ylab(displayTaxonName(taxon, taxaLevel)) +
                    geom_point() + 
                    scale_colour_manual(variable, values = myColorPalette) +
                    ggtitle( paste0("p-value: ", signif(pval, 5)) ) +
                    theme(plot.title = element_text(colour = ifelse(pval<=redPval, "red", "black"))) 
                variablePlotList[[taxon]] = gplot
            }
        }# taxon for loop
        
        
        # adjustedPvals
        padj.method="fdr"
        adjustedPvals = p.adjust(variablePvals, method = padj.method)
        adjPvalTable[,variable] = adjustedPvals
        
        # add row to summary table
        pvalCountTable = rbind(pvalCountTable, 
                               c(
                                   variable=variable,
                                   testType=testType,
                                   nVals=sum(!is.na(xAxis)),
                                   levels=paste0("min: ", min(xAxis, na.rm=T), "; mean: ", mean(xAxis, na.rm=T), "; median: ", median(xAxis, na.rm=T), "; max: ", max(xAxis, na.rm=T)),
                                   testWithPvalue=sum(!is.na(variablePvals)),
                                   rawPvalSig=sum(variablePvals <= redPval, na.rm=T)
                               ) 
        )
        
        
        if (doPlot){
            ord = order(variablePvals)
            if(saveToFile) {
                pdfname = paste0(fileNameBase, "_plots_", variable, ".pdf")
                pdf(pdfname)
                message("Saving file: ", pdfname)
            }
            for (i in ord){
                annotColor = ifelse(adjustedPvals[i] <= redPval, "red", "black")
                annotText = paste0(padj.method, "-adjusted p-value: ", signif(adjustedPvals[i], 5))
                show(
                    variablePlotList[[i]] +
                        annotate(geom="text", label=annotText, col=annotColor,
                                 x=Inf, y=Inf, hjust="inward", vjust="inward")
                )
            }
            if(saveToFile) {
                dev.off()
            }
        }
        
    }# variable for-loop
    
    if (saveToFile){
        # save summary table
        table.file = paste0(fileNameBase, "_summary.txt")
        message("Saving file: ", table.file)
        write.table(file=table.file, pvalCountTable, sep="\t", quote=F, row.names = F)
    }
    
    return(pvalTable)
}
```

```{r runTaxaVsNumericTest}
suppressWarnings({
    suppressMessages({
        peAtT1 = TaxaVsNumericTest(data %>% filter(TIMEPOINT == "T1"), 
                                   taxa, diffables, doPlot=T, testType = "pearson",
                                   fileNameBase = file.path(outDir, "T1.pearson"))
        keAtT1 = TaxaVsNumericTest(data %>% filter(TIMEPOINT == "T1"), 
                                   taxa, diffables, doPlot=F, testType = "kendall")
        peAtT2 = TaxaVsNumericTest(data %>% filter(TIMEPOINT == "T2"), 
                                   taxa, diffables, doPlot=F, testType = "pearson")
        keAtT2 = TaxaVsNumericTest(data %>% filter(TIMEPOINT == "T2"), 
                                   taxa, diffables, doPlot=F, testType = "kendall")
        peAtDif1 = TaxaVsNumericTest(diffData1, taxa, deltas, doPlot=F, testType = "pearson")
        keAtDif1 = TaxaVsNumericTest(diffData1, taxa, deltas, doPlot=F, testType = "kendall")
        peAtDif2 = TaxaVsNumericTest(diffData2, taxa, deltas, doPlot=F, testType = "pearson")
        keAtDif2 = TaxaVsNumericTest(diffData2, taxa, deltas, doPlot=F, testType = "kendall")
    })
})
```

# Test Summaries

### Combine test results

Merge to make one large heatmap.

Merge the categorical values.
```{r}
appendName <- function(df, suffix){
    names(df) = paste0(names(df), suffix)
    return(df)
}
mergeList = list(
    anovaAtT1 = appendName(anovaAtT1, ".HCT1.anova"),
    anovaAtT2 = appendName(anovaAtT2, ".HCT2.anova"),
    anovaAtHC = appendName(anovaAtHC, ".HC.anova"),
    acrossTimeTable = appendName(acrossTimeTable, ".T1T2.anova"),
    acrossTimeTable.acute = appendName(acrossTimeTable.acute, ".T1T2.Acute.anova"),
    acrossTimeTable.ceed = appendName(acrossTimeTable.ceed, ".T1T2.Ceed.anova"),
    acrossTimeTable.fargo = appendName(acrossTimeTable.fargo, ".T1T2.Fargo.anova"),
    kwAcrossTime = appendName(kwAcrossTime, ".T1T2.Kruskal"),
    kwAtT1 = appendName(kwAtT1, ".HCT1.Kruskal"),
    kwAtT2 = appendName(kwAtT2, ".HCT2.Kruskal"),
    kwAtHC = appendName(kwAtHC, ".HC.Kruskal")
    )

# merge
m.all.1 = data.frame(Row.names=taxa)
for (df in mergeList){
    m.all.1 = merge(m.all.1, df, by.x="Row.names", by.y=0)
}

# move row names
row.names(m.all.1) = m.all.1$Row.names
m.all.1 = m.all.1 %>% select(-Row.names)

# sort columns
m.all.1 = m.all.1[, order(names(m.all.1))]
# order groups of columns
m.all.1 = m.all.1 %>% select(starts_with("TIMEPOINT"),
                             starts_with("HCvsAN"),
                             starts_with("SUBTYPE.AN"),
                             starts_with("ANOREXIA"),
                             starts_with("LOCATION"),
                             everything())
```


Merge the numeric feature test results.
```{r}
appendName <- function(df, suffix){
    names(df) = paste0(names(df), suffix)
    return(df)
}
mergeList.numeric = list(
    peAtT1 = appendName(peAtT1, ".T1.pearson"),
    keAtT1 = appendName(keAtT1, ".T1.kendall"),
    peAtT2 = appendName(peAtT2, ".T2.pearson"),
    keAtT2 = appendName(keAtT2, ".T2.kendall"),
    peAtDif1 = appendName(peAtDif1, ".T1.pearson"),
    keAtDif1 = appendName(keAtDif1, ".T1.kendall"),
    peAtDif2 = appendName(peAtDif2, ".T2.pearson"),
    keAtDif2 = appendName(keAtDif2, ".T2.kendall")
    )

# merge
m.all.2 = data.frame(Row.names=taxa)
for (df in mergeList.numeric){
    m.all.2 = merge(m.all.2, df, by.x="Row.names", by.y=0)
}

# move row names
row.names(m.all.2) = m.all.2$Row.names
m.all.2 = m.all.2 %>% select(-Row.names)

# sort columns
m.all.2 = m.all.2[, order(names(m.all.2))]

# order groups of columns
m.all.2 = m.all.2 %>% select(starts_with("Weight_kg"),
                             starts_with("BMI"),
                             starts_with("PSS"),
                             starts_with("STAI"),
                             everything())
```


Merge into a single results table
```{r}
m.all = merge(m.all.1, m.all.2, by=0)

# move row names
row.names(m.all) = m.all$Row.names
m.all = m.all %>% select(-Row.names)

# orderRows = order(apply(m.all, MARGIN = 1, function(row){
#     sum(row < 0.05, na.rm=T)
#     }), decreasing = T)
orderRows = order(-log10(m.all$ANOREXIA.HCT1.anova), decreasing = T)
m.all = m.all[ orderRows, ]
```

### Save tables

Save p-values. Keep the row order specified above, but add the full names of taxa as an extra column.
```{r}
# row.names(nameKey) = nameKey$shortName
# m.all.labeled = cbind(taxon.originalName=nameKey[row.names(m.all),"originalName"], 
#                       taxon=row.names(m.all),
#                       m.all)
# 
# write.table(m.all.labeled, file = file.path(outDir, paste0("bigList_pvalues_raw_", taxaLevel, ".txt")),
#             row.names=F, quote=F, sep="\t")
```

Adjust p-values within each column, and save that table.  Add on the full taxon names, and keep row order.
```{r}
# m.all.adj = apply(m.all, MARGIN = 2, FUN=p.adjust, method="fdr")
m.all.adj = m.all
for( test in names(m.all)){
    m.all.adj[,test] = p.adjust(m.all[,test], method = "fdr")
}

# row.names(nameKey) = nameKey$shortName
# m.all.adj.labeled = cbind(taxon.originalName=nameKey[row.names(m.all.adj),"originalName"], 
#                       taxon=row.names(m.all.adj),
#                       m.all.adj)
m.all.adj.labeled = cbind(taxon=row.names(m.all.adj),
                      m.all.adj)

write.table(m.all.adj.labeled, file = file.path(outDir, paste0("bigList_pvalues_fdr-adjusted_", taxaLevel, ".txt")),
            row.names=F, quote=F, sep="\t")
```

Build a summary table for each variable tested:

  * How many tests have non-NA p-values?
  * How many raw p-values are significant?
  * How many are significant after adjusting for the number of taxa.
```{r}
summary = data.frame(testName="TOTAL",
                     attemptedTests=NA,
                     testWithPvalue=NA,
                     rawPvalSig=NA,
                     adjPvalSig_nTaxa=NA,
                     adjPvalSig_nTests=NA)

signifP = 0.05
for (testName in names(m.all)){
    # add row to summary table
    variablePvals = m.all[,testName]
    adjustedPvals = p.adjust(variablePvals, method="fdr")
    very.adjustedPvals = p.adjust(variablePvals, method="fdr", n = nrow(m.all) * ncol(m.all))
    summary = rbind(summary, 
                           c(
                               testName=testName,
                               attemptedTests=length(variablePvals),
                               testWithPvalue=sum(!is.na(variablePvals)),
                               rawPvalSig=sum(variablePvals <= signifP, na.rm=T),
                               adjPvalSig_nTaxa=sum(adjustedPvals <= signifP, na.rm=T),
                               adjPvalSig_nTests=sum(very.adjustedPvals <= signifP, na.rm=T)
                           ) 
    )
}
summary = summary %>% filter(testName != "TOTAL")

write.table(summary, file = file.path(outDir, paste0("bigList_summary_", taxaLevel, ".txt")),
            row.names=F, quote=F, sep="\t")
```


### pvalue heat map

Show the p-value table as a heat map.
```{r pvalHeat}
shortPathwayNames = function(longName){
  medName = gsub("[^ ]*: ", "", longName)
  isTooLong = nchar(medName) > 50
  if (isTooLong){
    trim1 = substr(medName, 0, 50)
    words = strsplit(trim1, split=" ")[[1]]
    lessWords = words[1:(length(words)-1)]
    trim2 = paste(c(lessWords, "..."), collapse = " ")
    medName = trim2
  }
  return(medName)
}

caption.about.names = "Pathway names were shortened. The pathway id was removed leaving the description.
Numbers were appened because the truncated names are not all unique.
Row order here (bottom to top) matches the row order in the corresponding p-value table."


pvalTableToHeatMap = function(pvalTable, midpointPval=0.05, bluePvalue=0.05, redPvalue=0.00001, title="-log10(p-values)", displayName.FUN=shortPathwayNames){
    # pvalTable - data frame of pvalues, with columns corresponding to variables (metadata) and named rows for taxa.
    # bluePval - pvalues worse than this will all be shown in blue.
    # midpointPval - these will be shown in white, as a middle point between the blue and red gradients.
    # all p-values better than this will be shown in solid red.
    pvalTable[pvalTable < redPvalue] = redPvalue
    pvalTable[pvalTable > bluePvalue] = bluePvalue
    
    taxon = sapply( row.names(pvalTable), displayName.FUN )
    if (length(unique(taxon)) < nrow(pvalTable)){
        taxon = paste0(taxon, "[", 1:nrow(pvalTable), "]")
    }
    row.names(pvalTable) = taxon
    
    longTab = -log10(pvalTable) %>% 
        mutate(taxon= row.names(pvalTable) ) %>%
        pivot_longer(cols=-taxon, names_to = "var", values_to="pval")
    longTab$var = factor(x=as.character(longTab$var), levels=names(pvalTable))
    longTab$taxon = factor(x=as.character(longTab$taxon), levels=row.names(pvalTable))
    ggplot(longTab, aes(x=var, y=taxon, fill=pval)) +
      geom_tile(col="gray") +
      scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                           midpoint = -log10(midpointPval),
                           limit = c(-log10(bluePvalue),-log10(redPvalue)),
                           space = "Lab",
                           name=paste0("p-value threshold: ", midpointPval, "\n\n-log10(p-value)"),
                           na.value = gray(.9)) +
      theme_minimal()+ # minimal theme
      ggtitle(title) +
      xlab("") +
      ylab(taxaLevel) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                       size = 8, hjust = 1)) +
      coord_fixed() +
      labs(caption=caption.about.names)
}
```


Heat-map for categorical feature p-values.
```{r}
pvalHeatMap.categories.file = file.path(outDir, paste0("heatmap_categoricalFeatures_raw-p-values_", taxaLevel, ".pdf"))
pdf(file=pvalHeatMap.categories.file,
    height = 3 + nrow(m.all.1) / 6,
    width = 5 + ncol(m.all.1) / 6)
pvalTableToHeatMap(m.all.1, title="catogrical-features raw p-values")
dev.off()
```

Heat-map for numerical feature p-values.
```{r}
pvalHeatMap.numeric.file = file.path(outDir, paste0("heatmap_numericFeatures_raw-p-values_", taxaLevel, ".pdf"))
pdf(file=pvalHeatMap.numeric.file,
    height = 3 + nrow(m.all.2) / 6,
    width = 5 + ncol(m.all.2) / 6)
pvalTableToHeatMap(m.all.2, title="numeric-features raw p-values")
dev.off()
```


Save a single large heatmap with all raw p-values
```{r}
pvalHeatMap.big.file = file.path(outDir, paste0("bigList_heatmap_raw-p-values_", taxaLevel, ".pdf"))
pdf(file=pvalHeatMap.big.file,
    height = 3 + nrow(m.all) / 6,
    width = 5 + ncol(m.all) / 6)
pvalTableToHeatMap(m.all, title="the big heatmap - raw p-values")
dev.off()
```

Save a single large heatmap with all ajdusted p-values
```{r}
pvalHeatMap.big.adj.file = file.path(outDir, paste0("bigList_heatmap_fdr.adj-p-values_", taxaLevel, ".pdf"))
pdf(file=pvalHeatMap.big.adj.file,
    height = 3 + nrow(m.all) / 6,
    width = 5 + ncol(m.all) / 6)
pvalTableToHeatMap(m.all.adj, title="the big heatmap - fdr-adjusted p-values")
dev.off()
```

### p-value histograms

Draw p-value histograms for all variables.
```{r}
pvalHist = function(pvalTable, title="", pdfName=title, saveToFolder=outDir, taxalevel=taxaLevel){
    longTab = cbind(taxon=row.names(pvalTable), pvalTable) %>% 
        pivot_longer(cols=-taxon, names_to = "variable", values_to = "value")
    # show(suppressMessages(
    #     ggplot(longTab, aes(x=value, fill=variable)) +
    #         geom_histogram() +
    #         ggtitle(paste0(title,"(", taxalevel, ")"))
    # ))
    if ( !is.null(saveToFolder) ){
        pdfName = file.path(outDir, paste0(title, "_pval-hist-", taxalevel, ".pdf"))
        message("Saving file: ", pdfName)
        pdf(pdfName, width=3, height = 3)
        for (column in names(pvalTable)){
            show(suppressMessages(
                ggplot(data = data.frame(x=pvalTable[,column]), aes(x=x)) +
                    geom_histogram(fill="dodgerblue") +
                    xlab("p-values") +
                    ggtitle(column)  
            ))
        }
        dev.off()
    }
    
}
```


```{r}
pvalHist(m.all, "bigList")
```

```{r}
sessionInfo()
```
